---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vespadbImportR

<!-- badges: start -->
[![Project Status: WIP - Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip) 
[![R-CMD-check](https://github.com/inbo/vespadbImportR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/inbo/vespadbImportR/actions/workflows/R-CMD-check.yaml) 
[![Codecov test coverage](https://codecov.io/gh/inbo/vespadbImportR/graph/badge.svg)](https://app.codecov.io/gh/inbo/vespadbImportR)
<!-- badges: end -->

## What is vespa-db

vespa-db is a database created by [vespawatch](https://vespawatch.be/en/) to make *Vespa velutina* **nest** observations available in as near as real time as possible.

The platform contains observations from different sources, but currently only new observations are added from [waarnemingen.be)[<https://waarnemingen.be/>]. In order for an observation to be included in vespa-db, it needs to have the following properties:

-   It has to currently be **identified** to *Vespa velutina* or a subspiecies thereof.
-   It has to have the activity: **Nest** (no observations of individuals are included)
-   It has to be **validated** by a validator, which means it has to have a photo. Automatically validated (based on location and likelyhood) observations are not included.

## How did this package come to be

This package was initially written by @PietrH in the context of importing historical data into vespa-db. For matching and validation purposes it was convenient to wrap the api into functions so imports and matching could be quickly tested. This original code resides in the private repository: inbo/vespaR.

Later on a need arose to query the vespa-db api again, and the original code was moved here for more broad reuse.

## The API
For an overview of all available API endpoints (certainly more than are wrapped here), see: https://db.vespawatch.be/swagger/.

## Installation

You can install the development version of vespadbImportR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("inbo/vespadbImportR")
```

## Examples

There are two options to get data out of vespa-db. The fastest way is to fetch the most recent export file. This file is generated every day at around 4AM Brussels time. 

```{r get export, }
library(vespadbImportR)
get_vespadb_export()
```

Alternativly you can also send queries to the database direclty and get live data back. This is much slower. 

```{r database query}
get_vespadb_obs(
  municipality_id = 32,
  min_observation_datetime = "2025-07-01T00:00:00",
  max_observation_datetime = "2025-07-25T00:00:00"
)
```
In this example we fetch data for the first 25 days of July 2025, for the municipality of Leuven.

To get an overview of the available parameters that you can use to query our observation data, see `get_observation_parameters()`.

```{r show available obs params}
get_observation_parameters()
```

You can also get an overview of the municipality codes, or query a municipality code directly with the function `municipalities`:

```{r demo municipalities}
municipalities("Leuven")
municipalities(c("Hasselt", "Leuven"))
```
```{r demo municipalities overview}
municipalities()
```



